---
title: "Wine Quality Analysis"
author: "David Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8, fig.height = 6
)
library(tidyverse)
library(GGally)
library(corrplot)
library(factoextra)
````

# 1. Project Overview

This analysis downloads the UCI Wine Quality datasets (red & white), merges them, explores key chemical and quality variables, performs PCA, and fits a regression model predicting wine quality.

# 2. Data Acquisition & Preparation

```{r data-load}
# URLs for the datasets
url_red   <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv"
url_white <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv"

# Column specification
wine_spec <- cols(
  `fixed acidity`        = col_double(),
  `volatile acidity`     = col_double(),
  `citric acid`          = col_double(),
  `residual sugar`       = col_double(),
  chlorides               = col_double(),
  `free sulfur dioxide`  = col_double(),
  `total sulfur dioxide` = col_double(),
  density                 = col_double(),
  pH                      = col_double(),
  sulphates               = col_double(),
  alcohol                 = col_double(),
  quality                 = col_integer()
)

# Read red and white
red   <- read_delim(url_red,   delim=";", col_types=wine_spec, locale=locale(decimal_mark=".")) %>%
  mutate(type="red")
white <- read_delim(url_white, delim=";", col_types=wine_spec, locale=locale(decimal_mark=".")) %>%
  mutate(type="white")

# Combine and clean
wine <- bind_rows(red, white) %>%
  mutate(
    type    = factor(type),
    quality = factor(quality, ordered=TRUE)
  )

cat("Total observations:", nrow(wine), "\n")
cat("Counts by type:\n"); print(table(wine$type))
```

# 3. Exploratory Visualizations

## 3.1 Wine Quality Counts by Type

```{r plot-quality-counts}
p1 <- ggplot(wine, aes(quality, fill=type)) +
  geom_bar(position="dodge") +
  labs(title="Wine Quality Counts by Type",
       x="Quality Score", y="Count") +
  theme_minimal()
p1
```

## 3.2 Alcohol Content vs. Quality

```{r plot-alcohol-vs-quality}
p2 <- ggplot(wine, aes(alcohol, quality, colour=type)) +
  geom_jitter(width=0, height=0.2, alpha=0.5) +
  labs(title="Alcohol Content vs. Quality",
       x="Alcohol (%)", y="Quality Score") +
  theme_minimal()
p2
```

## 3.3 Chemical Attributes by Quality

```{r plot-boxplots}
attrs <- c("pH","residual sugar","citric acid","sulphates")
p3 <- wine %>%
  pivot_longer(all_of(attrs), names_to="attribute", values_to="value") %>%
  ggplot(aes(quality, value, fill=quality)) +
  geom_boxplot() +
  facet_wrap(~attribute, scales="free_y") +
  labs(title="Chemical Attributes by Wine Quality") +
  theme_minimal() +
  theme(legend.position="none")
p3
```

## 3.4 Correlation Matrix of Numeric Attributes

```{r plot-corr, fig.width=7, fig.height=7}
nums <- wine %>% select_if(is.numeric)
corr <- cor(nums, use="pairwise.complete.obs")
corrplot(corr, method="color", type="upper",
         tl.col="black", tl.cex=0.8,
         title="Correlation Matrix of Wine Attributes")
```

# 4. Principal Component Analysis

## 4.1 Scree Plot

```{r pca-scree}
scaled <- scale(nums)
res.pca <- prcomp(scaled, center=TRUE, scale.=FALSE)
fviz_screeplot(res.pca, addlabels=TRUE, title="PCA: Variance Explained")
```

## 4.2 PCA Biplot

```{r pca-biplot}
fviz_pca_biplot(res.pca, repel=TRUE, title="PCA Biplot: Wines")
```

# 5. Regression Modeling

We treat quality as numeric and predict it using key variables.

```{r regression}
wine$num_quality <- as.numeric(as.character(wine$quality))
mod <- lm(num_quality ~ alcohol + sulphates + pH + type, data=wine)
cat("\n=== Regression Summary ===\n")
print(summary(mod))
```

## 5.1 Residuals vs. Fitted

```{r plot-residuals, fig.width=7, fig.height=5}
plot(mod, which=1)
```

## 5.2 Fitted vs. Actual

```{r plot-fitted-vs-actual}
wine$fitted <- fitted(mod)
p4 <- ggplot(wine, aes(fitted, num_quality, colour=type)) +
  geom_jitter(alpha=0.4, height=0.1) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  labs(title="Fitted vs. Actual Quality",
       x="Fitted Quality", y="Actual Quality") +
  theme_minimal()
p4
```

# 6. Conclusions

* **Quality distribution** differs by type: white wines skew slightly higher.
* **Alcohol content** shows a positive trend with quality.
* **PCA** reveals that the first two components capture the bulk of variance.
* **Regression** indicates that alcohol, sulphates, pH, and type significantly predict quality, though R² is modest.

# 7. References

* Cortez, Paulo, et al. “Modeling wine preferences by data mining from physicochemical properties.” *Decision Support Systems*, 47.4 (2009): 547–553.
* UCI Machine Learning Repository: Wine Quality Data Sets.
